---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pivot

*{pivot}* is an a package to pivot inside `dplyr::summarize()` calls. 
It's experimental, not thoroughly tested, etc...

Pivoting operations are grouped operations, although they are
traditionally treated differently, this package explores this idea .

The goal is to provide 2 functions, `widen()` and `lengthen()`, which do most of
what `pivot_wider()` and `pivot_longer()` do. Their key characteristics are : 

* They're used in a `dplyr::summarize()` call, which allows one to combine them with other aggregations
* The new names are on the left side of `=` or `:=`, the right side describes the values,
  which is what we're used to
* Very few arguments are needed (often just one), so it's very compact

The examples below take every example from `?tidyr::pivot_longer` and `?tidyr::pivot_wider`
and translates it into *{pivot}* syntax, right below the comented original code.

It has only been tested on these examples, please report if you find issues.

## Installation

You can install the released version of pivot from [CRAN](https://CRAN.R-project.org) with:

``` r
remotes::install_github("pivot")
```

## widen

`widen()` calls have the form `widen(..., .fill = NULL)`.

glue strings describe the names of the columns to create, values are an expression
of the column to take values from. 


```{r, eval = FALSE}
widen(
  "{col_to_take_new_names_from_1}" = col_to_take_values_from_1,
  "foo_{toupper(col_to_take_new_names_from_2)}" = col_to_take_values_from_2,
  .fill =  0
)

widen(
  "{col_to_take_new_names_from}" = mean(col_to_take_values_from)
)
```

## lengthen

`lengthen()` calls have the form `lengthen(expr, drop_na = FALSE)`.

On the left side, names of new columns are given, ending by the column containing the
values. 

On the right side a regular expression describes what columns should be
considered, and how to parse their names into the key columns.

`_val_` is a special value to handle multiple value columns.

```{r, eval = FALSE}
lengthen(
  c("key", "value") = "^myprefix"
)

# use groups for several key columns
lengthen(
  c("key1", "key2", "value") = "^(.*)_(.*)$"
)

# use named capture for a different order
lengthen(
  c("key1", "key2", "value") = "^(?<key2>.*)_(.*)$"
)

# use '_val_' to get several value columns
lengthen(
  c("key", "_val_") = "^(?<_val_>.*)_(.*)$"
)
```


## Examples from `?pivot_wider`

original code is commented right before our translation

```{r}
suppressPackageStartupMessages(library(tidyverse, warn.conflicts = FALSE))
library(pivot)

fish_encounters

# fish_encounters %>%
#   pivot_wider(names_from = station, values_from = seen)
fish_encounters %>%
  group_by(fish) %>%
  summarize(
    widen("{station}"= seen), 
    .groups = "drop")

# Fill in missing values
# fish_encounters %>%
#   pivot_wider(names_from = station, values_from = seen, values_fill = 0)
fish_encounters %>%
  group_by(fish) %>%
  summarize(
    widen("{station}" = seen, .fill = 0), 
    .groups = "drop")

# Generate column names from multiple variables
us_rent_income

# us_rent_income %>%
#   pivot_wider(names_from = variable, values_from = c(estimate, moe))
us_rent_income %>%
  group_by(GEOID, NAME) %>%
  summarize(widen("estimate_{variable}" = estimate,
                  "moe_{variable}" = moe), 
            .groups ="drop")


# When there are multiple `names_from` or `values_from`, you can use
# use `names_sep` or `names_glue` to control the output variable names
# us_rent_income %>%
#   pivot_wider(
#     names_from = variable,
#     names_sep = ".",
#     values_from = c(estimate, moe)
#   )
us_rent_income %>%
  group_by(GEOID, NAME) %>%
  summarize(widen("estimate.{variable}" = estimate,
                  "moe.{variable}" = moe), 
            .groups ="drop")

# us_rent_income %>%
#   pivot_wider(
#     names_from = variable,
#     names_glue = "{variable}_{.value}",
#     values_from = c(estimate, moe)
#   )
us_rent_income %>%
  group_by(GEOID, NAME) %>%
  summarize(widen("{variable}_estimate" = estimate,
                  "{variable}_moe" = moe), 
            .groups ="drop")

# Can perform aggregation with values_fn
warpbreaks <- as_tibble(warpbreaks[c("wool", "tension", "breaks")])
warpbreaks

# warpbreaks %>%
#   pivot_wider(
#     names_from = wool,
#     values_from = breaks,
#     values_fn = mean
#   )
warpbreaks %>%
  group_by(tension) %>%
  summarize(
    widen("{wool}" = mean(breaks)),
    .groups = "drop"
  )
```


## Examples from `?pivot_longer`

```{r}
# Simplest case where column names are character data
relig_income

# relig_income %>%
#   pivot_longer(-religion, names_to = "income", values_to = "count")
relig_income %>%
  group_by(religion) %>%
  summarize(
    lengthen(c("income", "count") := ".*"),
    .groups = "drop")

# Slightly more complex case where columns have common prefix,
# and missing missings are structural so should be dropped.
billboard

# billboard %>%
#  pivot_longer(
#    cols = starts_with("wk"),
#    names_to = "week",
#    names_prefix = "wk",
#    values_to = "rank",
#    values_drop_na = TRUE
#  )
billboard %>%
  group_by(artist, track, date.entered) %>%
  summarize(
    lengthen(c("week", "rank") := "^wk(.*?)$", drop_na = TRUE),
    .groups = "drop")

# Multiple variables stored in column names
# who %>% pivot_longer(
#   cols = new_sp_m014:newrel_f65,
#   names_to = c("diagnosis", "gender", "age"),
#   names_pattern = "new_?(.*)_(.)(.*)",
#   values_to = "count"
# )

who %>%
  group_by(country, iso2, iso3, year) %>%
  summarize(lengthen(
    c("diagnosis", "gender", "age", "count") := "^new_?(.*)_(.)(.*)"
  )) 

# Multiple observations per row
anscombe
# anscombe %>%
#  pivot_longer(everything(),
#    names_to = c(".value", "set"),
#    names_pattern = "(.)(.)"
#  )

anscombe %>%
  as_tibble() %>%
  summarize(lengthen(
    c("set", "_val_") := "(?<_val_>.)(.)"
  ))
```

